---
title: "Antibiotic Resistant Bacteria per state"
date: "March 15, 2026"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_float:
            collapsed: true
        number_sections: false
        code_download: true
        df_print: kable
        code_folding: show
        mode: selfcontained
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

[Return to Homepage](../index.html)

# **Introduction**
Antibiotics have been in use since WWII to treat infections that were once thought to be incurable. They have revolutionized medicine and saved countless lives, and will continue to save countless more. However, there is a major drawback to the use of antibiotics. Over the years of usage of antibiotics, doctors have over prescribed, and patients have not finished taking doses as instructed, leading to resistance. There are hundreds, maybe even more, species of bacteria that have developed resistance to our everyday use of antibiotics, and some that have become resistant to our stronger last resort antibiotics. In this research, we will look at the data for a common resistant bacterial strain commonly found in healthcare settings, as well as looking at resistance found in other species. We will also talk about how we can reduce the amount of resistance occurring, as well as potential other options.    

# **Background**
Antibiotic-resistant bacteria are not specific to just one state; as we will see, there is resistance in every single state. Not only do we have bacteria that are resistant to our oldest forms of antibiotics, like penicillin, which was first discovered by Alexander Fleming when he left bacteria out in his lab while on vacation. We are also seeing resistance in some of our more potent antibiotics, like Vancomycin, Carbapenem, Daptomycin, and others. A lot of our treatments for bacteria now need a combination of different antibiotics just to rid patients of infection. 

There are a few major classes of antibiotics that I will go over here. The first of these is beta-lactams, which include penicillin, which was mentioned earlier. Beta-lactams affect a specific part of the bacterial cell wall called peptidoglcan preventing the wall from correctly coming together. Bacteria evade this with beta-lactamases to help degrade the beta-lactam ring in these particular antibiotics. Other types of antibiotics that inhibit the cell wall are Cephalosporins, Carbapenems, and Vancomycin. Bacteria have found ways to be resistant to all of these antibiotics. 
The second type of antibiotics are those that inhibit protein synthesis, usually by targeting a part of the ribosomes. This is usually one of the ribosomal sub units. Typical antibiotics in this class include Aminoglycosides, Tetracycline, Macrolides, Oxazolidinones, and Streptogramins. 

The third type are ones that inhibit the synthesis of nucleic acids. Each of the antibiotics does a different function in this category, but the usual ones include Sulfa drugs, Quinolones, and Rifampin. For all three types of antibiotics listed here bacteria have developed ways to resist by modifying, destroying antibiotics using enzymes, making the cell less permeable, the use of efflux pumps, or by passing particular process. 
In this study we will specifically look at MRSA (Methicillin-resistant Staphylococcus aureus) which is found in hospitals and care facilities around the world. These facilities are breeding grounds for antibiotic resistance because of the congregation of sick individuals. 

# **Data**
```{r, echo = FALSE, results = 'hide'}
# load all necessary libraries 
library(tidyverse)
library(ggplot2)
library(maps)
library(RColorBrewer)

# reading in MRSA data set 
MRSA = read.csv("C:/Users/kaipa/Downloads/Kaipauole.github.io/final_project/Standardized-Infection-Ratios-by-State-Map/Standardized-Infection-Ratios-by-State-Map.csv")

```
I found data already created and cleaned up in the format I desired. As stated earlier for simplicity sake, I will just be exploring one type of resistant bacteria strain (MRSA). The data will be used to make a Choropleth map where the different colors will represent the concentration of bacterial resistance for this particular strain. I will go into more detain on this later. Here I will show a portion of the data table found (Link?) 

```{r}
# shows the original data structure 
str(MRSA)

# shows a portion of our updated data set 
head(MRSA)

```

As noted before, the data was collected across all 50 states and some of the territories (56 rows) with 13 columns. I will go over the ones important for us here. I will not be including the territories in this data: 

- **state**
 - All 50 states and territories.
  
  
- **HAI**
 -Stands for hospital acquired infection.
 
  
- **Event.year**
 -When the data was acquired (2022).
 
  
- **Observed.infections**
 -Number of infections found in that state or territory (MRSA).
 
# **Exploratory Analysis**
## Infections per state 
Here we see the data on a Choropleth map, using the different colors we can see which states have the highest rate of MRSA infections in a hospital setting: 
```{r}
# raw map used
par(mar = c(0, 0, 0, 0))
map('state', fill = TRUE, col = 'grey', border = 'black', lwd = 0.5)


# map with data
MRSA$State = tolower(MRSA$State)
MRSA$Observed.Infections = as.numeric(MRSA$Observed.Infections)
map_df = map_data("state")

MRSA_map = MRSA %>%
  mutate(bin = cut(Observed.Infections,
      breaks = c(-Inf, 125, 250, 375, 500, 625, 750, Inf),
      labels = c("0-125", "126-250", "251-375", "376-500", "501-625", "626-750", "750+"),
      include.lowest = TRUE))
map_df = left_join(map_df, MRSA_map,
                    by = c("region" = "State"))
ggplot(map_df, aes(long, lat, group = group, fill = bin)) +
  geom_polygon(color = "black") +
  scale_fill_brewer(
    palette = "Reds",
    name = "Observed Infections",
    direction = 1)+
  theme_void()
```

```
As you see from the map, most of the central states have lower MRSA observed infections. The east cost, California, Texas, and Florida have a higher observed infetions probably because of the higher populations in those areas. 

# **Disscussion**
This data just shows one strain of bacteria that is common in hospitals. As precription rates have gone up as well as misuse of antibiotics there has been increase in these super bugs. There has been promising research done on other tatics to reduce the use of antibiotic resistance. One such is the use of bacteriophages to target and kill resistant bacteria. There has even been cases where phage therapy has resulted in antibiotic resistance loss due to phage gene integration in the bacterial genome. 

# **Conclusion** 
As the rate of misuse and overperscription of antibiotics rises, the rate in antibiotic resistance will also climb. Because the develpment of new antibiotics is expensive and resistance is easy to develop there has been limited research in developing new antibiotics. However, promising research in other methods like phage therapy has shown promsing results and hopefuly will be used with antibiotic someday. In the mean time all we can do is use prescribed anitbiotics as directed, and finish the entire course effectivly killing all the bacteria and preventing futher resitance. 

# **Citations**
  United States | A.R. & Patient Safety Portal. (n.d.). Retrieved April 12, 2026, from https://cdc-earwig.datawheel.us/profile/geography/united-states?hai-select-sir=haiMRSA&phenotype-select-ar-by-hai=arPhenotype21&phenotype-select-ar-by-state=arPhenotype21&saar-age-select=age3&tabsection-182=1


[Return to Homepage](../index.html)
