---
title: "Antibiotic Resistant Bacteria per state"
date: "March 15, 2026"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_float:
            collapsed: true
        number_sections: false
        code_download: true
        df_print: kable
        code_folding: show
        mode: selfcontained
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

[Return to Homepage](../index.html)

# **Introduction**
Intro: Antibiotics have been in use since WWII to treat infections that were once thought to be incurable. They have revolutionized medicine and saved countless lives, and will continue to save countless more. However, there is a major drawback to the use of antibiotics. Over the years of usage of antibiotics, doctors have over prescribed, and patients have not finished taking doses as instructed, leading to resistance. There are hundreds, maybe even more, species of bacteria that have developed resistance to our everyday use of antibiotics, and some that have become resistant to our stronger last resort antibiotics. In this research, we will look at the data for a common resistant bacterial strain commonly found in healthcare settings, as well as looking at resistance found in other species. We will also talk about how we can reduce the amount of resistance occurring, as well as potential other options.    

# **Background**
Antibiotic-resistant bacteria are not specific to just one state; as we will see, there is resistance in every single state. Not only do we have bacteria that are resistant to our oldest forms of antibiotics, like penicillin, which was first discovered by Alexander Fleming when he left bacteria out in his lab while on vacation. We are also seeing resistance in some of our more potent antibiotics, like Vancomycin, Carbapenem, Daptomycin, and others. A lot of our treatments for bacteria now need a combination of different antibiotics just to rid patients of infection. 

There are a few major classes of antibiotics that I will go over here. The first of these is beta-lactams, which include penicillin, which was mentioned earlier. Beta-lactams affect a specific part of the bacterial cell wall called peptidoglcan preventing the wall from correctly coming together. Bacteria evade this with beta-lactamases to help degrade the beta-lactam ring in these particular antibiotics. Other types of antibiotics that inhibit the cell wall are Cephalosporins, Carbapenems, and Vancomycin. Bacteria have found ways to be resistant to all of these antibiotics. 
The second type of antibiotics are those that inhibit protein synthesis, usually by targeting a part of the ribosomes. This is usually one of the ribosomal sub units. Typical antibiotics in this class include Aminoglycosides, Tetracycline, Macrolides, Oxazolidinones, and Streptogramins. 

The third type are ones that inhibit the synthesis of nucleic acids. Each of the antibiotics does a different function in this category, but the usual ones include Sulfa drugs, Quinolones, and Rifampin. For all three types of antibiotics listed here bacteria have developed ways to resist by modifying, destroying antibiotics using enzymes, making the cell less permeable, the use of efflux pumps, or by passing particular process. 
In this study we will specifically look at MRSA (Methicillin-resistant Staphylococcus aureus) which is found in hospitals and care facilities around the world. These facilities are breeding grounds for antibiotic resistance because of the congregation of sick individuals. 

# **Data**
```{r, echo = FALSE, results = 'hide'}
# load all necessary libraries 
library(tidyverse)
library(ggplot2)
library(maps)
library(RColorBrewer)

# reading in MRSA data set 
MRSA = read.csv("C:/Users/kaipa/Downloads/Kaipauole.github.io/final_project/Standardized-Infection-Ratios-by-State-Map/Standardized-Infection-Ratios-by-State-Map.csv")

```
As already described, this data set was retrieved from Kaggle and had already been **mostly** cleaned for analysis. However, there are some changes I wanted to make to the data structure, missing values/empty spaces, and superfluous characters found throughout the data set. The following code and outputs demonstrate the changes I've made to allow for smoother data analysis: 

```{r}
# shows the original data structure 
str(MRSA)

# shows a portion of our updated data set 
head(MRSA)

```

As noted before, the data was collected accross 231 countries (231 rows) along with 6 particular statistics associated with each country. The columns are described as follows: 

- **serial_number**
  - a unique but arbitrary ID given to each row of data to serve as a key column 
  
- **country** 
  - the country in question 
  
- **total_cases**
  - total number of cases of COVID-19 reported at that point in the country 
  
- **total_deaths**
  - total number of deaths from COVID-19 reported at that point in the country 
  
- **total_recovered** 
  - total number of partients who recovered from COVID-19 at that point in the country 
  
- **active_cases**
  - currently on-going cases reported at the moment the survey was conducted 
  
- **total_test_taken**
  - total number of COVID-19 tests administered and reported at that point in the country 
  
- **population**
  - total number of people living in the country at that point of the survey
  
# **Exploratory Analysis**
## Death Rate vs. Testing Rate 
First, we will be investigating two metrics that aren't given to us in the actual data set: 
```


# **Conclusion** 
All my thoughts and interpretations have been strewn across this entire document so there isn't much for me to share here other than the point that there is much to be learned from what might seem like even the dullest of data. Data does not need to be super complex for it to be analyzed. Creative thought process of how to look at the data and what types of predictions you might be able to make with said data is just as important to. COVID-19 offers great insight into how the world can handle large medical crisis and if they did handle it well. That can be said for many other forms of data so be on the look out and think creatively! You'll never know what you may find. I also didn't sleep all night in the process of writing this entire script so I'm calling it good now, thank you for reading! 

[Return to Homepage](../index.html)
